### 2024-04-24 parts/appendix.tex
reshaping appendix

\AP
The following
\cref{nat-rel-poly:def} is one of the equivalent definitions of \cite{CDTL23},
and is similar in shape to the \emph{finite counting automata} introduced by
\cite{SCHU62}.

\begin{definition}[$\Rel$-polyregular functions {\cite{CDTL23}}]
    \label{nat-rel-poly:def}
    Let $M$ be a finite monoid, $\mu \colon \Sigma^* \to M$
    be a morphism, $k \in \Nat$, and 
    $\pi \colon M^k \to \Rel$ be a function.
    The \intro{$\Rel$-polyregular function}
    $\pi^\dagger \colon \Sigma^* \to \Rel$
    is computed as follows:
    \begin{equation*}
        \pi^\dagger (w) \defined
        \sum_{w = u_1 \cdots u_k} \pi(\mu(u_1), \dots, \mu(u_k))
        \quad .
    \end{equation*}
    When $\pi$ has co-domain $\Nat$, the function $\pi^\dagger$
    is called \intro{$\Nat$-polyregular}.
    We call $\pi$ the \emph{production function} of $\pi^\dagger$.
    When the monoid $M$ is \kl(monoid){aperiodic}
    then
    the function $f$ is \intro{star-free $\Rel$-polyregular}
    (resp. \intro{star-free $\Nat$-polyregular}).
\end{definition}

\textbf{TODO: remove this example}
\begin{example}
    \label{size-of-word-nsf:ex}
    The map $f \colon w \mapsto \card{w} + 1$
    belongs to $\NSF$.
    \proofref{size-of-word-nsf:ex}
\end{example}

@@

\AP Let us briefly recall in the following lemmas alternative characterizations
of $\NPoly$ (resp. $\ZPoly$), some of which will be helpful in the upcoming
analysis of their \kl{commutative} counterpart, and others that provide a nicer
syntax to construct examples. These characterizations were previously known,
and can be found for instance in \cite{DOUE23,CDTL23}. For every language $L
\subseteq \Sigma^*$ we define $\ind{L} \colon \Sigma^* \to \set{0,1}$ as the
indicator function of $L$. Given an $\MSO$ (resp. $\FO$) formula with
first-order free variables $\varphi(\vec{x})$, we define
$\vcount{\varphi(\vec{x})}$ as the function from $\Sigma^*$ to $\Nat$ that maps
a word $w$ to the number of assignments $\nu \colon \vec{x} \to w$, such that
$w, \nu \models \varphi(\vec{x})$. If $X$ is a set of functions from $\Sigma^*$
to $\Rel$, and a subset $\mathbb{S}$ of $\Rel$, we define
$\intro*\Span{\mathbb{S}}(X)$ as the smallest subset of functions from
$\Sigma^* \to \Rel$ containing $X$ and closed under (finite) sums and
multiplication with elements $s \in \mathbb{S}$.

%Similar characterizations can be obtained for $\NSF$ and $\ZSF$, which correctly
%generalizes aperiodicity to functions as \cref{regular-language:ex} illustrates.
%\begin{example}
%    \label{regular-language:ex}
%    Let $\Sigma$ be a finite alphabet, and
%    $L \subseteq \Sigma^*$. Then,
%    $L$ is a regular language if and only if
%    $\ind{L} \colon \Sigma^* \to \set{0,1}$ is a
%    \kl{$\Nat$-polyregular function}.
%    Furthermore, $L$ is a star-free regular language
%    if and only if $\ind{L}$ is a
%    \kl{star-free $\Nat$-polyregular function}.
%\end{example}
%It has been proven in \cite{CDTL23,BOJA22} that the classes $\ZPoly[k]$ (resp.
%$\NPoly[k]$, $\ZSF[k]$, $\NSF[k]$) are also characterized by the maximal number
%of first order free variables used to construct the function as in
%\cref{npoly-counting-mso:item} of \cref{polynomial-rational-polyreg:fact}.

\begin{proofof}[size-of-word-nsf:ex]
    Let us define $M \defined (\set{1}, \times)$ which is 
    a finite \kl{aperiodic monoid}, $\mu \colon \Sigma^* \to M$
    defined by $\mu(w) \defined 1$, and
    $\pi \colon M^2 \to \Nat$
    that is the constant function equal to $1$.
    We check that for all $w \in \Sigma^*$:
    \begin{equation*}
        \pi^\dagger(w)
        =
        \sum_{uv = w} 1
        =
        \card{w} + 1
        = f(w)
        \quad . 
        \qedhere
    \end{equation*}
\end{proofof}

@@

\begin{proofof}[n-poly-combinatorics:lem]
    Let $\omega$ be an idempotent power for the finite monoid $M$,
    and
    $w \defined \alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i)$.
    Let $k \in \Nat$, and $\pi \colon M^k \topartial \Nat$ be such that
    $f = \pi^\dagger$. That is, 
    $f(w)$ is the sum over all factorizations of $w$
    into $k$ words $(w_1, \dots, w_k) \in \Sigma^*$
    of the value $\pi(\mu(w_1), \dots, \mu(w_k))$.

@@

\begin{proofof}[n-rat-correct:lem]
    Let $f$ be the \kl{commutative}
    \kl{$\Nat$-rational series}
    with domain defined as $\Sigma \defined \set{a_1, \dots, a_p}$
    that \kl{represents} $P$. Because $f$ has \kl{polynomial growth},
    it is in fact a \kl{$\Nat$-polyregular function}
    (\cref{polynomial-rational-polyreg:fact}).
    Using \cref{n-poly-combinatorics:lem},
    there exists a number $\omega \in \Nat$
    and polynomial $Q$ with \kl{non-negative}
    \kl{maximal monomials} such that
    for all $X_1, \dots, X_p \geq \omega$:
    \begin{equation*}
        f\left(
            \prod_{i = 1}^p a_i^{\omega X_i}
        \right)
        = Q(X_1, \dots, X_p)
        = P(\omega X_1, \dots, \omega X_p) 
        \quad .
    \end{equation*}
    We conclude that $P(\omega X_1, \dots, \omega X_p) = Q(X_1, \dots, X_p)$
    as polynomials,
    and in particular that
    the \kl{maximal monomials} of 
    $P$ are \kl{non-negative}.

@@

\begin{lemma}
    \label{derivation-simplifies:lemma}
    For all $P \in \Rel[\vec{X}]$ that are non-constant,
    for all $K \in \Nat$,
    $\MaximalMonomials(\Diff{K}{P}) \hoarele
    \MaximalMonomials(P)$.
    \proofref{derivation-simplifies:lemma}
\end{lemma}
\begin{proofof}[derivation-simplifies:lemma]
    Let $P \in \Rel[\vec{X}]$ and
    $K \in \Nat$.
    Leveraging 
    \cref{derivation-monomials:fact,discrete-deriv-linear:fact,maxi-monomials-submodular:fact}:
    \begin{align*}
        \MaximalMonomials\left(\Diff{K}{P}\right)
        &=
        \MaximalMonomials\left(\sum_{M \in \Monomials(P)} \Diff{K}{M}\right) \\
        &\hoareleq
        \bigcup_{M \in \Monomials(P)} 
        \MaximalMonomials(\Diff{K}{M}) \\
        &\hoarele
        \MaximalMonomials(P)
        \quad .
        \qedhere
    \end{align*}
\end{proofof}

@@

\begin{lemma}
    \label{n-poly-k-implies-wqo:lemma}
    Let $k \in \Nat$, and let $f \in \NPoly[k]$.
    Then, $(\Sigma^* \resleq{f}{k})$ is a \kl{well-quasi-ordering}.
    \proofref{n-poly-k-implies-wqo:lemma}
\end{lemma}
\begin{proofof}[n-poly-k-implies-wqo:lemma]
    Because $f \in \NPoly[k]$, there exists
    a tuple $\vec{x}$ of first order free variables,
    $\MSO$ formulas $\seqof{\psi_i(\vec{x})}{1 \leq i \leq n}$,
    and positive coefficients $\seqof{m_i}{1 \leq i \leq n}$,
    such that
    $f = \sum_{i = 1}^n m_i \times \vcount{\varphi_i(\vec{x})}$.

@@

    Let $\seqof{u_i}{i \in \Nat}$ be an infinite sequence of $\Sigma^*$.
    Without loss of generality, one can assume that for all $i \neq j$, $u_i
    \equiv_k u_j$, i.e., that the difference $\app{f}{u_i} - \app{f}{u_j}$
    belongs to $\ZPoly[k-1]$, since the latter has finite index. Thanks to the
    above remarks, there exists $i < j$ such that $\MSO^q(u_i) \leq
    \MSO^q(u_j)$. As a consequence, $g \defined \app{f}{u_j} - \app{f}{u_i} \in
    \NPoly$, and therefore $g \in \NPoly[k-1]$. We have proven that there
    exists $i < j$ such that $u_i \resleq{f}{k} u_j$.
\end{proofof}

>>>>

@@

@@

\AP In order to prove \cref{n-poly-combinatorics:lem}, we will rely on an
equivalent definition of \kl{$\Nat$-polyregular functions} that is more
combinatorial in nature, and is similar in shape to the \emph{finite counting
automata} introduced by \cite{SCHU62}.

\begin{lemma}[$\Rel$-polyregular functions {\cite{CDTL23}}]
    \label{nat-rel-poly:def}
    Let $M$ be a finite monoid, $\mu \colon \Sigma^* \to M$
    be a morphism, $k \in \Nat$, and 
    $\pi \colon M^k \to \Rel$ be a function.
    The following function 
    $\pi^\dagger \colon \Sigma^* \to \Rel$
    is \kl{$\Rel$-polyregular}:
    \begin{equation*}
        \pi^\dagger (w) \defined
        \sum_{w = u_1 \cdots u_k} \pi(\mu(u_1), \dots, \mu(u_k))
        \quad .
    \end{equation*}
    When $\pi$ has co-domain $\Nat$, the function $\pi^\dagger$
    is furthermore \kl{$\Nat$-polyregular}.
    We call $\pi$ the \emph{production function} of $\pi^\dagger$.
    When the monoid $M$ is \kl(monoid){aperiodic}
    then
    the function $f$ is \kl{star-free $\Rel$-polyregular}
    (resp. \kl{star-free $\Nat$-polyregular}).
\end{lemma}

In order to illustrate the above definition, let us provide an example of a
\kl{$\Nat$-polyregular function} computed using a finite monoid.

\begin{example}
    \label{size-of-word-nsf:ex}
    The map $f \colon w \mapsto \card{w} + 1$
    belongs to $\NSF$.
\end{example}
\begin{proof}
    Let us define $M \defined (\set{1}, \times)$ which is 
    a finite \kl{aperiodic monoid}, $\mu \colon \Sigma^* \to M$
    defined by $\mu(w) \defined 1$, and
    $\pi \colon M^2 \to \Nat$
    that is the constant function equal to $1$.
    We check that for all $w \in \Sigma^*$:
    \begin{equation*}
        \pi^\dagger(w)
        =
        \sum_{uv = w} 1
        =
        \card{w} + 1
        = f(w)
        \quad . 
        \qedhere
    \end{equation*}
\end{proof}

\begin{proofof}[n-poly-combinatorics:lem]
    We leverage the characterization of \kl{$\Nat$-polyregular functions}
    given in \cref{nat-rel-poly:def}.
    Let $\omega$ be an idempotent power for the finite monoid $M$,
    and
    $w \defined \alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i)$.
    Let $k \in \Nat$, and $\pi \colon M^k \topartial \Nat$ be such that
    $f = \pi^\dagger$. That is, 
    $f(w)$ is the sum over all factorizations of $w$
    into $k$ words $(w_1, \dots, w_k) \in \Sigma^*$
    of the value $\pi(\mu(w_1), \dots, \mu(w_k))$.

@@

\begin{proofof}[n-rat-correct:lem]
    Let $f$ be the \kl{commutative}
    \kl{$\Nat$-rational series}
    with domain defined as $\Sigma \defined \set{a_1, \dots, a_p}$
    that \kl{represents} $P$. Because $f$ has \kl{polynomial growth},
    it is in fact a \kl{$\Nat$-polyregular function}
    (\cite[Theorem 5.22]{DOUE23}).
    Using \cref{n-poly-combinatorics:lem},
    there exists a number $\omega \in \Nat$
    and polynomial $Q$ with \kl{non-negative}
    \kl{maximal monomials} such that
    for all $X_1, \dots, X_p \geq \omega$:
    \begin{equation*}
        f\left(
            \prod_{i = 1}^p a_i^{\omega X_i}
        \right)
        = Q(X_1, \dots, X_p)
        = P(\omega X_1, \dots, \omega X_p) 
        \quad .
    \end{equation*}
    We conclude that $P(\omega X_1, \dots, \omega X_p) = Q(X_1, \dots, X_p)$
    as polynomials,
    and in particular that
    the \kl{maximal monomials} of 
    $P$ are \kl{non-negative}.

@@

\begin{lemma}
    \label{derivation-simplifies:lemma}
    For all $P \in \Rel[\vec{X}]$ that are non-constant,
    for all $K \in \Nat$,
    $\MaximalMonomials(\Diff{K}{P}) \hoarele
    \MaximalMonomials(P)$.
\end{lemma}
\begin{proof}
    Let $P \in \Rel[\vec{X}]$ and
    $K \in \Nat$.
    Leveraging 
    \cref{derivation-monomials:fact,discrete-deriv-linear:fact,maxi-monomials-submodular:fact}:
    \begin{align*}
        \MaximalMonomials\left(\Diff{K}{P}\right)
        &=
        \MaximalMonomials\left(\sum_{M \in \Monomials(P)} \Diff{K}{M}\right) \\
        &\hoareleq
        \bigcup_{M \in \Monomials(P)} 
        \MaximalMonomials(\Diff{K}{M}) \\
        &\hoarele
        \MaximalMonomials(P)
        \quad .
        \qedhere
    \end{align*}
\end{proof}

@@

\begin{lemma}
    \label{n-poly-k-implies-wqo:lemma}
    Let $k \in \Nat$, and let $f \in \NPoly[k]$.
    Then, $(\Sigma^* \resleq{f}{k})$ is a \kl{well-quasi-ordering}.
\end{lemma}
\begin{proof}
    Because $f \in \NPoly[k]$, there exists
    a tuple $\vec{x}$ of first order free variables,
    $\MSO$ formulas $\seqof{\psi_i(\vec{x})}{1 \leq i \leq n}$,
    and positive coefficients $\seqof{m_i}{1 \leq i \leq n}$,
    such that
    $f = \sum_{i = 1}^n m_i \times \vcount{\varphi_i(\vec{x})}$.

@@

    Let $\seqof{u_i}{i \in \Nat}$ be an infinite sequence of $\Sigma^*$.
    Without loss of generality, one can assume that for all $i \neq j$, $u_i
    \equiv_k u_j$, i.e., that the difference $\app{f}{u_i} - \app{f}{u_j}$
    belongs to $\ZPoly[k-1]$, since the latter has finite index. Thanks to the
    above remarks, there exists $i < j$ such that $\MSO^q(u_i) \leq
    \MSO^q(u_j)$. As a consequence, $g \defined \app{f}{u_j} - \app{f}{u_i} \in
    \NPoly$, and therefore $g \in \NPoly[k-1]$. We have proven that there
    exists $i < j$ such that $u_i \resleq{f}{k} u_j$.
\end{proof}

### 2024-04-24 parts/prelim.tex
reshaping appendix

\AP The notion of \kl{$\mathcal{H}$-transducer} can be used to inductively
define the classes $\intro*\NPoly[k]$, $\intro*\NSF[k]$, $\intro*\ZPoly[k]$ and
$\intro*\ZSF[k]$ as follows \cite[Theorems 5.15 and 7.10]{DOUE23}: in the base
case, $\reintro*\NPoly[-1] = \reintro*\NSF[-1]  = \reintro*\ZSF[-1] =
\reintro*\ZPoly[-1]$ is the singleton containing the constant function equal to
$0$; and for $k \geq 0$, $\reintro*\ZPoly[k]$ is the set of functions
\kl{computed} by a \kl{$\reintro*\ZPoly[k-1]$-transducer}, and
$\reintro*\NPoly[k]$ is the set of functions \kl{computed} by an
\kl{$\reintro*\NPoly[k-1]$-transducer} \emph{with an $\Nat$-valued output
function $F$}, and similarly for $\reintro*\NSF[k]$ and $\reintro*\ZSF[k]$,
where the transducer is furthermore required to be \kl{counter-free}. Then,
$\reintro*\NPoly = \bigcup_{k \in \Nat} \reintro*\NPoly[k]$, and classes
$\reintro*\NSF$, $\reintro*\ZPoly$ and $\reintro*\ZSF$ are defined similarly.
In the rest of the paper, we will always assume that
\kl{$\reintro*\NPoly[k]$-transducers} and \kl{$\reintro*\NSF[k]$-transducers}
have an $\Nat$-valued output functions, making a clean separation between
$\NPoly$ and $\ZPoly$.

>>>>

\AP The notion of \kl{$\mathcal{H}$-transducer} can be used to inductively
define the classes $\intro*\NPoly[k]$, $\intro*\NSF[k]$, $\intro*\ZPoly[k]$ and
$\intro*\ZSF[k]$ as follows \cite[Theorems 5.15 and 7.10]{DOUE23}: in the base
case, $\reintro*\NPoly[-1] = \reintro*\NSF[-1]  = \reintro*\ZSF[-1] =
\reintro*\ZPoly[-1]$ is the singleton containing the constant function equal to
$0$; and for $k \geq 0$, $\reintro*\ZPoly[k]$ is the set of functions
\kl{computed} by a \kl{$\reintro*\ZPoly[k-1]$-transducer}, and
$\reintro*\NPoly[k]$ is the set of functions \kl{computed} by an
\kl{$\reintro*\NPoly[k-1]$-transducer} \emph{with an $\Nat$-valued output
function $F$}, and similarly for $\reintro*\NSF[k]$ and $\reintro*\ZSF[k]$,
where the transducer is furthermore required to be \kl{counter-free}. Then,
$\reintro*\NPoly = \bigcup_{k \in \Nat} \reintro*\NPoly[k]$, and classes
$\reintro*\NSF$, $\reintro*\ZPoly$ and $\reintro*\ZSF$ are defined analogously.
In the rest of the paper, we will always assume that
\kl{$\reintro*\NPoly[k]$-transducers} and \kl{$\reintro*\NSF[k]$-transducers}
have an $\Nat$-valued output functions, making a clean separation between
$\NPoly$ and $\ZPoly$.
